Narrative

### 2023-05-24 manuscript/manuscript.qmd
manuscript: added in-text-values for dynamic values

tables <- readRDS("tables.RDS")
library(kableExtra) #must be loaded to print saved kableextra tables

@@

Figure 1 show the absolute and relative bias in heterogeneity estimates for different levels of heterogeneity (measured in $\tau$) and meta-analytic models. Each column of Figure 1 corresponds to one nominal level of heterogeneity ($\tau$ $\epsilon$ {0, 0.1, 0.15, 0.2}) and each row to a different analysis (HV with Fisher $z$, HV with Pearson's $r$, and HS). The black unbroken lines indicate actual heterogeneity, which may differ from nominal heterogeneity because of truncation or in the case of Fisher $z$ because the comparable heterogeneity levels were slightly different (see methods). The different dashed lines in Figure 1 shows the $\hat{\tau}$ estimates for four different levels of average reliability across studies, and the x-axis shows the average (superpopulation) effect size. As can be seen, the bias is marginally worse for Fisher z than the other two methods (although Fisher z is on a different scale) but the results are approximately the same for all three methods. As such, we will focus in the remainder of the results on Pearson's r (row two).

@@

As expected, for zero heterogeneity (leftmost column Figure 1) we see an overestimate of heterogeneity that increases with effect size. There are two sources of positive bias in the absence of heterogeneity. First, and the only one that applies when $\mu = 0$, there is positive bias due to the truncation of negative heterogeneity estimates (Viectbauer, 2005). Second, there is bias due to the variance in reliability. For example, for $mu = 0.2$ (the closest value to the median correlation reported by Schäfer & Schwarz) and $\bar{R} = 0.8$ the bias for Pearson's $r$ is XX. For $\mu = 0.4$ (the upper median reported by Schäfer & Schwarz) and $\bar{R} = 0.8$ the bias is XX. For the highest average reliability (0.9) a somewhat less inclined slope can be seen. This is because the standard deviation for this reliability level is severely truncated when the reliability is bounded at 1. The increase in positive bias with effect size can also be seen in the presence of heterogeneity (all other columns). However, it is superseded by the negative bias.

For all heterogeneity levels in Figure 1 above zero we see an overall underestimate of heterogeneity due to unreliability in primary studies. This negative bias can be relatively severe, is worse with lower effect size and lower average reliability, and worsens in an absolute sense but improves in a relative sense as heterogeneity increases. For example, for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.1$, the bias is XX (XX%) for Pearson's r, for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.15$ the bias is XX (XX%), and for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.2$ the bias is XX (XX%).

To explore further under what conditions the positive and negative bias in heterogeneity estimates would reach an equilibrium, we added an analysis for several conditions with $\tau < 0.1$. We used the following values {0.02, 0.04, 0.06, 0.08}. Figure 2 presents the results of this analysis for Pearson's r. Because these lower heterogeneity levels can be expected to be affected by truncation to a larger extent (which results in a positive bias), we present results for four different numbers of studies in the meta-analysis (rows Figure 2, lower $k$ is associated with larger positive bias). The standard deviation in reliabilities is again 0.15. Figure 2 shows that even when only little heterogeneity is present, it will generally be underestimated. If $\mu < 0.3$ then imperfect reliability leads to a negative bias in heterogeneity estimates for all $\tau < 0.06$ in Pearson's $r$. Even for lower levels of heterogeneity, heterogeneity will generally be positively biased only when average effect size is large or the number of meta-analyzed studies is small.

>>>>

in_text <- readRDS("in-text-values.RDS")
tables <- readRDS("tables.RDS")
library(kableExtra) #must be loaded to print saved kableextra tables

@@

Figure 1 show the absolute and relative bias in heterogeneity estimates for different levels of heterogeneity (measured in $\tau$) and meta-analytic models. Each column of Figure 1 corresponds to one nominal level of heterogeneity ($\tau$ $\epsilon$ {0, 0.1, 0.15, 0.2}) and each row to a different analysis (HV with Fisher $z$, HV with Pearson's $r$, and HS). The black unbroken lines indicate actual heterogeneity, which may differ from nominal heterogeneity because of truncation or in the case of Fisher $z$ because the comparable heterogeneity levels were slightly different (see methods). The different dashed lines in Figure 1 shows the $\hat{\tau}$ estimates for four different levels of average reliability across studies, and the x-axis shows the average (superpopulation) effect size. As can be seen, the bias is marginally worse for Fisher $z$ than the other two methods (although Fisher $z$ is on a different scale) but the results are approximately the same for all three methods. As such, we will focus in the remainder of the results on Pearson's $r$ (row two).

@@

As expected, for zero heterogeneity (leftmost column Figure 1) we see an overestimate of heterogeneity that increases with effect size. Although Pearson's $r$ has a small positive bias due to the dependance between effect size and sampling variance (supplement XX), more generally there are two sources of positive bias in the absence of heterogeneity. First, and the only one that applies when $\mu = 0$, there is positive bias due to the truncation of negative heterogeneity estimates (Viechtbauer, 2005). Second, there is bias due to the variance in reliability. For example, for $mu = 0.2$ (the closest value to the median correlation reported by Schäfer & Schwarz) and $\bar{R} = 0.8$ the bias for Pearson's $r$ compared to true zero heterogeneity is `r in_text$bias_tau_zero$mu_0.2`. For $\mu = 0.4$ (the upper median reported by Schäfer & Schwarz) and $\bar{R} = 0.8$ the bias is `r in_text$bias_tau_zero$mu_0.4`. For the highest average reliability (0.9) a somewhat less inclined slope can be seen. This is because the standard deviation for this reliability level is severely truncated when the reliability is bounded at 1. The increase in positive bias with effect size can also be seen in the presence of heterogeneity (all other columns). However, it is superseded by the negative bias.

For all heterogeneity levels in Figure 1 above zero we see an overall underestimate of heterogeneity due to unreliability in primary studies. This negative bias can be relatively severe, is worse with lower effect size and lower average reliability, and worsens in an absolute sense but improves in a relative sense as heterogeneity increases. For example, for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.1$, the bias is `r in_text$bias_increasing_tau$tau_0.1` for Pearson's $r$, for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.15$ the bias is `r in_text$bias_increasing_tau$tau_0.15`, and for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.2$ the bias is `r in_text$bias_increasing_tau$tau_0.2`.

To explore further under what conditions the positive and negative bias in heterogeneity estimates would reach an equilibrium, we added an analysis for several conditions with $\tau < 0.1$. We used the following values {0.02, 0.04, 0.06, 0.08}. Figure 2 presents the results of this analysis for Pearson's $r$. Because these lower heterogeneity levels can be expected to be affected by truncation to a larger extent (which results in a positive bias), we present results for four different numbers of studies in the meta-analysis (rows Figure 2, lower $k$ is associated with larger positive bias). The standard deviation in reliabilities is again 0.15. Figure 2 shows that even when only little heterogeneity is present, it will generally be underestimated. If $\mu < 0.3$ then imperfect reliability leads to a negative bias in heterogeneity estimates for all $\tau < 0.06$ in Pearson's $r$. Even for lower levels of heterogeneity, heterogeneity will generally be positively biased only when average effect size is large or the number of meta-analyzed studies is small.
